\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Chapter 11}
\paragraph{1} M\&Ms are expected to have the following distribution:\\
\smallskip
{\centering
\begin{tabular}{r | c c c c c c }
Color & Blue & Brown & Green & Orange & Red & Yellow\\
Percent & 24\% & 14\% & 15\% & 20\% & 13\% & 14\% 
\end{tabular}
\par} 
\begin{enumalpha}
\item What is the minimum number of M\&Ms needed to do a valid goodness-of-fit test against the expected distribution. In other words, what is the sample size $n$ so that the smallest expected value is at least 5?\\
\medskip
\bt{Since red is the least frequent color in the expected distribution, we need to find a sample size so that there are at least 5 expected red M\&M's.}\\
\medskip
$\ds\bv{n \times 0.13 \ge 5, \qquad n \ge \frac 5 {0.13} = 38.46}$\\
\medskip
\bt{So, the sample size should be at least 39 M\&M's.}
\vspace{.5in}
\item Conduct a goodness-of-fit test of whether the distribution of M\&Ms is what is claimed by the company at a significance level of $\alpha=0.05$. Make sure to state the null and alternative hypotheses, and your conclusion in context of question.
\end{enumalpha}



\newpage
\paragraph{2} The Tortilla and Cheese Organization (TACO) thinks that preferences for types of tacos are the same for men and women. They conduct a survey and collect the following data (``taco\_preference.csv" on D2L):\\
\medskip
{\centering
\begin{tabular}{c | c  c c c}
\multicolumn{1}{c}{} & \multicolumn{4}{c}{\large Type of taco}\\
Gender & Beef & Pork & Chicken & Fish\\
\hline
Men & 105 & 34 & 56 & 27\\
Women & 83 & 29 & 75 & 35 \\
\end{tabular}
\par}
\bigskip
Test the claim the taco preference is the same for men and women at $\alpha = 0.05$ level of significance. Make sure to state the null and alternative hypotheses, and your conclusion in context of question.\\
\medskip
\bt{$\bv{H_0:}$ There is no association between gender and taco preference, or gender and taco preference are independent}\\
\bt{$\bv{H_a:}$ There is an association between gender and taco preference, or gender and taco preference are not independent}\\
\medskip
\bt{From StatCrunch Chi square test of independence: }$\bv{\chi^2 = 6.76, \, p = 0.08 > \alpha = 0.05}$\\
\bt{Fail to reject $\bv{H_0}$. There is no evidence that taco preference and gender are not independent.}


\vspace{.5in}
\paragraph{3} The file ``hair\_eye.csv" on D2L contains data on the hair color, eye color and gender for statistics students. Test whether there is an association between hair color and gender at significance level $\alpha=0.05$. Make sure to state the null and alternative hypotheses, and your conclusion in context of question. Is there an association between eye color and gender?\\
\medskip
\bt{$\bv{H_0:}$ There is no association between gender and hair color, or gender and hair are independent}\\
\bt{$\bv{H_a:}$ There is an association between gender and hair color, or gender and hair color are not independent}\\
\medskip
\bt{From StatCrunch chi square test of independence: }$\bv{\chi^2 = 7.99, \, p = 0.0461 < \alpha = 0.05}$\\
\bt{Reject $\bv{H_0}$. There is evidence that gender and hair color are associated.}



\end{flushleft}
\end{document}