\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Chapter 9}
\paragraph{1} A coffee shop is interested in the proportion of decaf coffee drinkers on Sunday and Monday mornings. The manager thinks they have a lower proportion of decaf drinkers on Monday. They examine a random sample of coffee orders and find that on Sunday 52 out of 156 orders are for decaffeinated coffee and on Monday 43 out 174 are decaf orders.
\begin{enumalpha}
\item What are the null and alternative hypotheses for a test on this claim? Is this a one-sided or two-sided test? Are the requirements for a hypothesis test satisfied?\\
\medskip
$\bv{H_0: p_s = p_m \quad \text{\bt{or}} \quad p_s - p_m = 0}$ \\
$\bv{H_a: p_s > p_m \quad \text{\bt{or}} \quad p_s - p_m > 0}$ \\
\bt{One-sided test}\\
\bt{All the requirements are satisfied.}
\vspace{.5in}
\item Conduct an hypothesis test at the $\alpha = 0.05$ level of significance. Be sure to state your conclusion in the context of the question.\\
\medskip
$\bv{z = 1.73, \, p = 0.0421 < \alpha = 0.05}$. \bt{Reject $\bv{H_0}$}.\\
\bt{There is evidence that the proportion of decaf coffee orders is lower on Mondays.}
\vspace{.5in}
\item Construct an appropriate confidence interval of the true difference of population proportions. What is your conclusion based on the confidence interval? Does it agree with the hypothesis test?\\
\medskip
\bt{90\% confidence interval: (0.004, 0.168)}\\
\bt{Because both lower and upper limits are positive, we can conclude, with 95\% certainty, that the true population difference in proportions is positive. Thus, this result agrees with the hypothesis test.}
\end{enumalpha}



\newpage
\paragraph{2} A manufacturer of flash drives wants to know if there is a difference in the reliability of their drives used in extreme conditions. A sample of 15 drives used in cold conditions ($< 32 $\textdegree F) had a mean lifespan of 41.9 months with a standard deviation of 6.3. A sample of 15 drives used in hot conditions ($> 99 $\textdegree F) had a mean lifespan of 38.4 months with a standard deviation of 5.9. Assume lifespans of flash drive are normally distributed.
\begin{enumalpha}
\item What are the null and alternative hypotheses for a test on this claim? Is this a one-sided or two-sided test? Are these independent or dependent samples? Are the requirements for a hypothesis test satisfied?\\
\medskip
$\bv{H_0: \mu_c = \mu_h \quad \text{\bt{or}} \quad \mu_c - \mu_h = 0}$ \\
$\bv{H_a: \mu_c \ne \mu_h \quad \text{\bt{or}} \quad \mu_c - \mu_h \ne 0}$ \\
\bt{Two-sided test, independent samples}\\
\bt{All the requirements are satisfied (populations are normally distributed).}
\vspace{.5in}
\item Conduct an hypothesis test at the $\alpha = 0.05$ level of significance. Be sure to state your conclusion in the context of the question.\\
\medskip
$\bv{t = 1.57, \, p = 0.1276 > \alpha =0.05}$\bt{. Fail to reject }$\bv{H_0.}$\\
\bt{There is not evidence the drives in extreme cold and extreme heat have different lifespans.}
\vspace{.5in}
\item Construct an appropriate confidence interval of the true difference of population means. What is your conclusion based on the confidence interval? Does it agree with the hypothesis test?\\
\medskip
\bt{95\% confidence interval: (-1.066, 8.066)}\\
\bt{Because zero is contained in the interval, we cannot conclude that the true population difference in mean lifespans is different than zero. Thus, this result agrees with the hypothesis test.}
\end{enumalpha}

\newpage
\paragraph{3} Researchers are interested in whether meditation can lower blood pressure in people that have high blood pressure. They conduct a study on 45 patients with high blood pressure (systolic blood pressure $> 20$), measuring their systolic blood pressure at baseline and after 30 minutes of meditation. The file ``meditation\_bp.csv" on D2L contains the data.
\begin{enumalpha}
\item What are the null and alternative hypotheses for a test on this claim? Is this a one-sided or two-sided test? Are these independent or dependent samples? Are the requirements for a hypothesis test satisfied?\\
\medskip
\bt{Let $\bv{\mu_d}$ be the population mean difference in systolic blood pressure (SBP after 30 minutes of meditation - baseline SBP).}\\
$\bv{H_0: \mu_d = 0}$ \\
$\bv{H_a: \mu_d < 0}$ \\
\bt{One-sided test, dependent or paired samples}\\
\bt{All the requirements are satisfied (sample size is $\bv{> 30}$).}
\vspace{.5in}\item Conduct an hypothesis test at the $\alpha = 0.01$ level of significance. Be sure to state your conclusion in the context of the question.\\
\medskip
$\bv{t = -6.34, \, p < 0.0001 < \alpha =0.01}$\bt{. Reject }$\bv{H_0.}$\\
\bt{There is evidence that systolic blood pressure is lower after 30 minutes of meditation.}
\vspace{.5in}
\item Construct an appropriate confidence interval of the true population mean of the differences. What is your conclusion based on the confidence interval? Does it agree with the hypothesis test?\\
\medskip
\bt{98\% confidence interval: (-3.283, -1.472)}\\
\bt{Because both lower and upper limits are negative, we can conclude, with 99\% certainty, that the true population mean of the change in systolic blood pressure between baseline and after 30 minutes of meditation (after meditation - baseline) is negative. Thus, this result agrees with the hypothesis test.}
\end{enumalpha}



\end{flushleft}
\end{document}