\documentclass{beamer}

\usepackage{shyne}

% Theme settings
\setbeamertemplate{navigation symbols}{}

\usetheme{Madrid}

\title[Short title]{Sample Analysis Presentation}
\author[M. Shyne]{\Large Michael Shyne}
\date{April, 2, 2018}

\begin{document}

\frame{\titlepage}

\begin{frame}
\frametitle{News Item}
\bt{\large Attending a Concert Every 2 Weeks Will Add 9 Years to Your Life, Study Finds}\\
\medskip
from: Digital Music News\\
\smallskip
(\url{https://www.digitalmusicnews.com/2018/03/28/music-concert-gigs-nine-years/})
\end{frame}

\begin{frame}{Statistics in article}
\begin{quote}
``O2 and Goldsmith University Associate Lecturer Patrick Fagan published the unexpected results \ldots" 
\end{quote}
\begin{quote}
``According to the study, experiencing a gig for just 20 minutes can result in a 21\% increase in feelings of well-being. Furthermore, the research found a direct link between “high levels of well-being and a lifespan increase of nine years” or more."
\end{quote}
{\centering
\bt{and}
\par}
\medskip
\begin{quote}
``Psychometric testing and heart-rate tests were given to the participants of the study. Results showed that the participants increased their feelings of self-worth (25\%), closeness to others (25\%) and mental stimulation (75\%) when they attended gigs."
\end{quote}
\end{frame}

\begin{frame}{Source of statistics}
\begin{itemize}
\item No title of paper published about study given.
\item A link in the story went to another news article about the study (\href{http://www.nme.com/news/music/going-to-gigs-can-help-you-live-longer-new-study-report-patrick-fagan-o2-2274410}{``Going to gigs can help you live longer, new study claims"})
\item That page has a link to another story on the website \href{https://news.o2.co.uk/press-release/science-says-gig-going-can-help-you-live-longer-and-increases-wellbeing/}{``O2 The Blue"}
\item On the O2 site, we find out that:
\begin{itemize}
\item ``O2 commissioned scientific study reveals the power of live music on the human body and mind"
\item ``The O2 is the world’s most popular music and entertainment venue" and ``O2 customers have access to tickets to over 5,000 shows in over 350 venues across the UK each year via Priority Tickets"
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Potential problems}
\begin{itemize}
\item The study was paid for by a company that would profit from people going to more concerts, indicating a possible source of bias.
\item The study does not appear to have undergone any peer-review process.
\item The description of study methods used is vague.
\item The stats used have unclear meanings (what does ``21\% more feelings of wellbeing" mean?).
\item The logical path of the claim, that going to concerts increases wellbeing $\to$ people who have wellbeing live longer (from other research), is tenuous at best.
\item There is a confusion between correlation, though the Digital Music News story acknowledges this.

\end{itemize}
\end{frame}

\end{document}