\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Chapter 10}
\paragraph{1} The file ``Galton-mother-daughter.csv" contains a subset 50 subjects from Galton's mother/daughter height data.
\begin{enumalpha}
\item Create a scatterplot of the data. Does there appear to be a linear relationship between mother's heights and daughter's heights?
\vspace{1.75in}
\item Conduct a correlation hypothesis test at $\alpha = 0.05$ significance level. If there is significant correlation, how would you describe the strength of the correlation?
\vspace{1.75in}
\item Find the estimated regression line for the relationship between mother's heights (predictor variable) and daughter's heights (response variable)? Is the slope significantly different than zero?
\vspace{1.75in}
\item What is the best predicted daughter's height for a mother that is 56 inches tall? Is it appropriate to make such a prediction?
\end{enumalpha}



\newpage
\paragraph{2} The Prime Minister of Freedonia wishes to improve the economic strength of his country. He instructs his finance minister to study the relationship between tacos consumed and Freedonia's GDP. The results by month for last year are in the file ``Freedonia.csv" on D2L. The data represents tacos consumed per capita and GDP in billlions of Freedonia's national currency (Fr\$).
\begin{enumalpha}
\item Create a scatterplot of the data. Does there appear to be a linear relationship between tacos consumed and Freedonia's GDP?
\vspace{1.6in}
\item Conduct a correlation hypothesis test at $\alpha = 0.05$ significance level. If there is significant correlation, how would you describe the strength of the correlation? Would it be worthwhile to also test the correlation with GDP measured in US dollars?
\vspace{1.6in}
\item Find the estimated regression line for the relationship between tacos consumed (predictor variable) and GDP (response variable)? Is the slope significantly different than zero?
\vspace{1.6in}
\item What is the best predicted GDP for a month in which 9.5 tacos are consumed per capita? Is it appropriate to make such a prediction?
\end{enumalpha}

\newpage
\paragraph{3} The file ``MCA\_scores\_17.csv" on D2L contains average math MCA scores for 11th graders in 2017 by MN public school district, as well as percentage of 11th graders receiving free lunches in the district. Districts with missing data and charter schools are excluded.
\begin{enumalpha}
\item Create a scatterplot of the data. Does there appear to be a linear relationship between percentage of students receiving free lunch and average MCA scores?
\vspace{1.75in}
\item Conduct a correlation hypothesis test at $\alpha = 0.01$ significance level. If there is significant correlation, how would you describe the strength of the correlation?
\vspace{1.75in}
\item Find the estimated regression line for the relationship between percentage of students receiving free lunch (predictor variable) and average MCA scores (response variable)? Is the slope significantly different than zero?
\vspace{1.75in}
\item What is the best predicted average MCA score for a district that has 45\% of 11th grade students receiving free lunch? Is it appropriate to make such a prediction?
\end{enumalpha}



\end{flushleft}
\end{document}