\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Week 5}
\paragraph{1} A small town does a census of pet ownership. They find the following distribution of number of dog per household\ldots\\
{\centering
\begin{tabular}{c | c}
Number of dogs & Probability\\
\hline
0 & 0.394\\
1 & 0.342\\
2 & 0.16\\
3 & 0.067\\
4 & 0.036\\
5 & 0.001\\
\end{tabular} \par}
\begin{enumalpha}
\item Is this a probability distribution? What is the expected number of dogs per household? What is the standard deviation? What is the variance?
\vspace{3in}
\item What is the probability of a household having at most one dog? Is this unusual? What is the probability of having four or more dogs? Is this unusual?
\end{enumalpha}



\newpage
\paragraph{2} The best player on the Metro State basketball team successfully makes 85\% of her free throws. During a typical game she attempts 15 free throws. (If she attempts a different number of free throws, it is not a typical game.)
\begin{enumalpha}
\item Is the number of free throws she makes in a typical game a proper binomial random variable? What are the values of $n$, $p$ and $q$?

\vspace{1.8in}
\item What is the expected number of free throws made per game? What is the standard deviation?

\vspace{1.8in}
\item What is the probability she makes all her free throws in a typical game? Is that unusual? What is the probability she makes at least 12 free throws? Less than 12?

\vspace{1.8in}
\item What are unusual numbers of free throws made per game?
\end{enumalpha}

\newpage
\paragraph{3} Our taco restaurant, which is open 12 hours a day, get an average of 720 customers a day. For staffing purposes, they are interested in understanding how many customers they might get between 1 and 2 PM. They want to use a Poisson random variable of customers per hour.
\begin{enumalpha}
\item What is the rate of customers per hour ($\lambda$)? If $X \sim Pois(\lambda)$ is the random variable to be used, what is the expected value of $X$? What is the standard deviation?

\vspace{1.8in}
\item What is the probability that the restaurant will get 70 or more customers in an hour? What is the probability they will get less than 55? 

\vspace{1.8in}
\item What are unusually high or low numbers of customers per hour?

\vspace{1.8in}
\item What is wrong with using this method to predict customers during the hour between 1 and 2 pm?
\end{enumalpha}



\end{flushleft}
\end{document}