\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Chapter 7}
\paragraph{1} Suppose is in known that 45\% of the general population believes that correlation implies causation.
\begin{enumalpha}
\item A survey taken of 60 students after completing a statistics course finds that 32\% of them believe that correlation implies causation. Find a 95\% confidence interval of the population proportion of students who have completed a statistics course who believe that correlation implies causation. Write it in both interval notation, $(L, U)$ or $L < p < U$, and in $\hat p \pm ME$ notation.
\vspace{2in}
\item Is the proportion of students who have completed a statistics course who believe that correlation implies causation different than the general population at a $\alpha = 0.05$ significance level? Did the statistics courses cause the difference, if it exists?

\vspace{2.25in}
\item If we wanted to known the population proportion of students who have completed a statistics course who believe that correlation implies causation within plus/minus 3\% with 90\% confidence, what sample size would be needed for a survey? Calculate with both unknown sample proportion and with proportion found in the earlier survey in part (a).
\end{enumalpha}



\newpage
\paragraph{2} Suppose summer daily maximum temperatures in Minnesota are known to be normally distributed with a standard deviation 6.5 \textdegree F. 
\begin{enumalpha}
\item A random sample of 16 summer day maximum temperatures has a sample mean of  71.1 \textdegree F. What is 90\% confidence interval for the population mean maximum temperature? Write it in both interval notation, $(L, U)$ or $L < \mu < U$, and in $\bar x \pm ME$ notation. 
\vspace{2.5in}
\item Suppose the file ``max\_temps.csv" represents a random sample of max temperatures in August. What is a 99\% confidence interval based on this sample? (Don't have to write both forms.) Is the mean maximum August temperature different than the mean max summer temperature (as found in part (a)) at a significance level of $\alpha = 0.01$?
\vspace{2.5in}
\item If we wanted to find the mean summer maximum temperature within plus/minus 1.5 \textdegree F with 95\% confidence, how many days would need to be sampled?
\end{enumalpha}



\end{flushleft}
\end{document}